---
title: "Gut transplants analysis"
author: "Rachel Yost, Alison Fowler, Lynn Adler"
output: html_document
---

# Libraries 
```{r}
library(lattice)
library(psych)
library(car)
library(AER)
library(MASS)
library(corrplot)
library(AICcmodavg)
library(plyr)
library(dplyr)
library(tidyr)
library(lme4)
library(mgcv)
library(scatterplot3d)
library(bbmle)
library(emmeans)
library(ggplot2)
library(tidyverse)
library(ggalt)
library(multcomp)
library(coxme)  
library(survival)  
library(DHARMa)
library(aod)  
library(pscl) 
library(writexl)
library(glmmTMB)
library(survminer)
library(fitdistrplus)
```

# Set up data 

```{r}
# setwd(" ")
feces <- read_csv("Yost_fecal_screens.csv")
bees <- read_csv("Yost_gut_counts.csv")
```

```{r}
# make factors 
bees$Treatment <- as.factor(bees$Treatment)
bees$Colony <- as.factor(bees$Colony)
bees$Crithidia_date <- as.factor(bees$Crithidia_date)

# remove NAs
bees_noNAs <- bees %>% filter(Wing_mm != is.na(Wing_mm))

# sample sizes
summary(bees_noNAs$Crithidia_date)
summary(bees_noNAs$Treatment)
```

# Gut infection data 

## Summaries 
```{r}
summary(bees$Colony)
summary(bees$Treatment)
summary(bees$Crithidia_date)

bees %>% group_by(Colony, Treatment) %>% summarize(length(Treatment))
bees %>% group_by(Crithidia_date, Treatment) %>% summarize(length(Treatment))
bees %>% group_by(Donor1_start, Treatment) %>% summarize(length(Treatment))
bees %>% group_by(Gut1_date, Treatment) %>% summarize(length(Treatment))

hist(bees$Count)
```

## Model selection

```{r}
count_mod1 <- glmmTMB(Count ~ Treatment * Crithidia_date * Colony, data = bees, family = nbinom2())

count_mod2 <- glmmTMB(Count ~ Treatment + Wing_mm + (1|Crithidia_date) + (1|Colony), data = bees, family = nbinom2())

count_mod3 <- glmmTMB(Count ~ Treatment + Wing_mm + (1|Crithidia_date), data = bees, family = nbinom2())

count_mod4 <- glmmTMB(Count ~ Treatment + Wing_mm + (1|Colony), data = bees, family = nbinom2())

count_mod5 <- glmmTMB(Count ~ Treatment + Wing_mm, data = bees, family = nbinom2())

count_mod6 <- glmmTMB(Count ~ Treatment, data = bees, family = nbinom2())

count_mod7 <- glmmTMB(Count ~ Treatment + Wing_mm + Crithidia_date, data = bees, family = nbinom2())

count_mod8 <- glmmTMB(Count ~ Treatment + Wing_mm + Colony, data = bees, family = nbinom2())

count_mod9 <- glmmTMB(Count ~ Treatment + Wing_mm + Colony + Crithidia_date, data = bees, family = nbinom2())

count_mod10 <- glmmTMB(Count ~ Treatment * Wing_mm, data = bees, family = nbinom2())

count_mod11 <- glmmTMB(Count ~ Treatment + Wing_mm*Crithidia_date, data = bees, family = nbinom2())

AICtab(count_mod1, count_mod2, count_mod3, count_mod4, count_mod5, 
       count_mod6, count_mod7, count_mod8, count_mod9, count_mod10, count_mod11)

# 5 is best but we have some problems with residuals 
simulateResiduals(count_mod5, plot = T)
# 11 was next best (delta AIC 0.2) - residuals look better, no red lines 
simulateResiduals(count_mod11, plot = T)
```

```{r}
# Add dispersion fomrula to mod 5 
count_mod5.1 <- glmmTMB(Count ~ Treatment + Wing_mm, data = bees, dispformula = ~Treatment*Wing_mm, family = nbinom2())
count_mod5.2 <- glmmTMB(Count ~ Treatment + Wing_mm, data = bees, family = nbinom2()) # no disp, same as mod 5 
count_mod5.3 <- glmmTMB(Count ~ Treatment + Wing_mm, data = bees, dispformula = ~Treatment + Wing_mm, family = nbinom2())
count_mod5.4 <- glmmTMB(Count ~ Treatment * Wing_mm, data = bees, dispformula = ~Treatment * Wing_mm, family = nbinom2())

AICtab(count_mod5.1,count_mod5.2, count_mod5.3, count_mod5.4, count_mod11)
#             dAIC df
# count_mod5.1 0.0  7 
# count_mod5.3 1.0  6 
# count_mod5.4 1.5  8 
# count_mod5.2 5.8  4 
# count_mod11  6.1  6 

simulateResiduals(count_mod5.1, plot = T)  
simulateResiduals(count_mod5.2, plot = T)
simulateResiduals(count_mod5.3, plot = T)
simulateResiduals(count_mod5.4, plot = T)
```

```{r}
summary(count_mod5.1)
Anova(count_mod5.1)

summary(count_mod5.3)
Anova(count_mod5.3)

summary(count_mod5.4)
Anova(count_mod5.4, type = "III")
```

```{r}
# try adding dispersion formula to model 11 
count_mod11.2 <- glmmTMB(Count ~ Treatment + Wing_mm*Crithidia_date, dispformula = ~Wing_mm*Crithidia_date, data = bees, family = nbinom2())

count_mod11.3 <- glmmTMB(Count ~ Treatment + Wing_mm + Crithidia_date, dispformula = ~Wing_mm*Crithidia_date, data = bees, family = nbinom2())

# remove date, go back to mod 5 but keep date in dispersion formula 
count_mod5.5 <- glmmTMB(Count ~ Treatment + Wing_mm, dispformula = ~Wing_mm*Crithidia_date, data = bees, family = nbinom2())
simulateResiduals(count_mod5.5, plot = T)  

count_mod5.6 <- glmmTMB(Count ~ Treatment + Wing_mm + (1|Crithidia_date), dispformula = ~Wing_mm*Crithidia_date, data = bees, family = nbinom2())

count_mod5.7 <- glmmTMB(Count ~ Treatment + Wing_mm + (1|Crithidia_date), dispformula = ~Wing_mm + Crithidia_date, data = bees, family = nbinom2())
simulateResiduals(count_mod5.7, plot = T)
summary(count_mod5.7)

count_mod5.8 <- glmmTMB(Count ~ Treatment + Wing_mm, dispformula = ~Wing_mm + Crithidia_date, data = bees, family = nbinom2())
simulateResiduals(count_mod5.8, plot = T) # best so far 
Anova(count_mod5.8)

count_mod11.4 <- glmmTMB(Count ~ Treatment + Wing_mm + Crithidia_date, dispformula = ~Wing_mm + Crithidia_date, data = bees, family = nbinom2())
simulateResiduals(count_mod11.4, plot = T) # also good 
summary(count_mod11.4)

AICtab(count_mod5.1,count_mod5.2, count_mod5, count_mod5.3, count_mod5.4, count_mod11, count_mod11.2, count_mod11.3, count_mod5.5, count_mod5.6, count_mod5.7, count_mod5.8, count_mod11.4)

#               dAIC df
# count_mod5.8   0.0 6 
# count_mod5.5   2.0 7 
# count_mod11.4  2.0 7 
# count_mod5.7   2.0 7 
# count_mod11.3  4.0 8 
# count_mod5.6   4.0 8 
# count_mod11.2  4.9 9 
# count_mod5.1   5.3 7 
# count_mod5.3   6.3 6 
# count_mod5.4   6.8 8 
# count_mod5.2  11.1 4 
# count_mod5    11.1 4 
# count_mod11   11.4 6 

# check zero inflation
count_mod5.8.zi <- glmmTMB(Count ~ Treatment + Wing_mm, dispformula = ~Wing_mm + Crithidia_date, ziformula = ~Crithidia_date, data = bees, family = nbinom2())

AICtab(count_mod5.8, count_mod5.8.zi)
#                 dAIC df
# count_mod5.8    0    6 
# count_mod5.8.zi 3    8 
simulateResiduals(count_mod5.8.zi, plot = T)
summary(count_mod5.8.zi)
```

MODEL TO USE: count_mod5.8 

```{r}
simulateResiduals(count_mod5.8, plot = T) # looks great 
Anova(count_mod5.8)
# 
#            Chisq Df Pr(>Chisq)  
# Treatment 0.0694  1    0.79227  
# Wing_mm   2.9622  1    0.08523 .
```

## Plot gut infections 
```{r}
means <- emmeans(count_mod5.8, ~ Treatment, type = "response")
counts.means.to.plot<-as.data.frame(summary(means))
counts.means.to.plot$upperSE <- counts.means.to.plot$response + counts.means.to.plot$SE
counts.means.to.plot$lowerSE <- counts.means.to.plot$response - counts.means.to.plot$SE
counts.means.to.plot
```

```{r}
gut_barplot<- 
  ggplot(counts.means.to.plot,aes(x=Treatment, y=response, ymin=lowerSE, ymax=upperSE)) + 
  theme_classic() +
  geom_bar(stat="identity", aes(fill = Treatment), color = "black") +
  scale_fill_manual(values=c("sienna4", "goldenrod1"),
                    labels = c("buckwheat","sunflower")) +
  ylab(expression(paste(italic("Crithidia"), " cells in 0.02 μL of gut"))) +
  xlab("\n Treatment") +
  geom_errorbar(width=0.05, position=position_dodge(width = 0.075))+
  theme(text = element_text(size=18),
        legend.position = "none") +
  scale_x_discrete(labels = c("Buckwheat", "Sunflower")) +
  ylim(0, 22) + 
  geom_text(x = 1.5, y = 20, label = expression(paste("Treatment: ", italic("P"), "= 0.79")), size = 5)

gut_barplot
```

## Gut count presence/absence
### Make presence/absence column

```{r}
# remove NAs 
bees_noNAs <- bees %>% 
  filter(Wing_mm != is.na(Wing_mm))
  
# make a new column with 0 or 1
bees_noNAs$Cells_P <- NA

for(i in 1:length(bees_noNAs$Count)){
  if(bees_noNAs$Count[i] > 0) {
    bees_noNAs$Cells_P[i] <- 1
  } else {
    bees_noNAs$Cells_P[i] <- 0
  }
} 
```

### Model presence/absence 

```{r}
gut_PA_mod <- glmer(Cells_P ~ Treatment + Wing_mm + (1|Crithidia_date) + (1|Colony), 
                      family = binomial(link = "logit"), data = bees_noNAs)

gut_PA_mod2 <- glmer(Cells_P ~ Treatment + Wing_mm + (1|Colony), 
                      family = binomial(link = "logit"), data = bees_noNAs)

gut_PA_mod3 <- glmer(Cells_P ~ Treatment + Wing_mm + (1|Crithidia_date), 
                      family = binomial(link = "logit"), data = bees_noNAs)

gut_PA_mod4 <- glm(Cells_P ~ Treatment + Wing_mm, 
                      family = binomial(link = "logit"), data = bees_noNAs)

AICtab(gut_PA_mod, gut_PA_mod2, gut_PA_mod3, gut_PA_mod4) # 2 best 

simulateResiduals(gut_PA_mod2, plot = T) # looks good 
Anova(gut_PA_mod2)
```

### Plot presence/absence 
```{r}
gut.PA.means <- emmeans(gut_PA_mod2, ~ Treatment, type = "response")
gut.PA.means.to.plot<-as.data.frame(summary(gut.PA.means))
gut.PA.means.to.plot$upperSE <- gut.PA.means.to.plot$prob + gut.PA.means.to.plot$SE
gut.PA.means.to.plot$lowerSE <- gut.PA.means.to.plot$prob - gut.PA.means.to.plot$SE
gut.PA.means.to.plot
```

```{r}
gut_PA_barplot <- 
  ggplot(gut.PA.means.to.plot,aes(x=Treatment, y=prob, ymin=lowerSE, ymax=upperSE)) + 
  theme_classic() +
  geom_bar(stat="identity", aes(fill = Treatment), color = "black") +
  scale_fill_manual(values=c("sienna4", "goldenrod1"),
                    labels = c("buckwheat","sunflower")) +
  ylab("Probability of cells in 0.02 μL gut") +
  xlab("\n Treatment") +
  geom_errorbar(width=0.05, position=position_dodge(width = 0.075))+
  theme(text = element_text(size=18),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  scale_x_discrete(labels = c("Buckwheat", "Sunflower")) +
  ylim(0, 1.1) + 
  geom_text(x = 1.5, y = 1, label = expression(paste("Treatment: ", italic("P"), "= 0.44")), size = 5) 

gut_PA_barplot
```

# Fecal screens 

## Get set up

```{r}
bees_sub <- bees[,1:10]
feces_dates2 <- merge(feces, bees_sub, by = "Bee_ID")
feces_dates2$Treatment.x <- as.factor(feces_dates2$Treatment.x)

# rename these columns 
colnames(feces_dates2)[8] <- "Cells_per_ul"
colnames(feces_dates2)[5] <- "Feces_line_length_mm"
colnames(feces_dates2)[10] <- "Feces_conc"
colnames(feces_dates2)[4] <- "Cells_presence"

# rename bw and sf 
levels(feces_dates2$Treatment.x) <- c("Buckwheat", "Sunflower")

# add column for volume 
feces_dates2 <- feces_dates2 %>% mutate(fecal_vol = ((Feces_line_length_mm*25)/65))
```

## Summaries

```{r}
feces_dates2 %>% group_by(Colony, Treatment.x) %>% summarize(length(Treatment.x))
feces_dates2 %>% group_by(Treatment.x) %>% summarize(length(Treatment.x))
```

## Models 
 
### Model bee size by colony
```{r}
hist(feces_dates2$Wing_mm)
wing_mod <- lm(Wing_mm ~ Colony, data = feces_dates2)
simulateResiduals(wing_mod, plot = T)
Anova(wing_mod)
```

### Model fecal volume 
```{r}
hist(feces_dates2$fecal_vol)

# with date as fixed effect 
fecal_vol_mod <- lm(fecal_vol ~ Treatment.x + Wing_mm + Crithidia_date, data = feces_dates2)
simulateResiduals(fecal_vol_mod, plot = T)  
vif(fecal_vol_mod)
Anova(fecal_vol_mod)

# with colony as fixed effect 
fecal_vol_mod2 <- lm(fecal_vol ~ Treatment.x + Wing_mm + Colony, data = feces_dates2)
simulateResiduals(fecal_vol_mod2, plot = T)
Anova(fecal_vol_mod2)

fecal_vol_mod3 <- lm(fecal_vol ~ Colony, data = feces_dates2)
simulateResiduals(fecal_vol_mod3, plot = T)
Anova(fecal_vol_mod3)

fecal_vol_mod4 <- lm(fecal_vol ~ Colony + Wing_mm + Treatment.x, data = feces_dates2)
simulateResiduals(fecal_vol_mod4, plot = T)
Anova(fecal_vol_mod4)

fecal_vol_mod5 <- lm(fecal_vol ~ Wing_mm, data = feces_dates2)
simulateResiduals(fecal_vol_mod5, plot = T)
Anova(fecal_vol_mod5)
```

### Model cells per ul feces

#### Remove NAs 
```{r}
hist(feces_dates2$Cells_per_ul, breaks= 100)

# remove NAs
feces_dates2_noNAs <- feces_dates2 %>% 
  filter(Cells_presence != is.na(Cells_presence))
```

#### Model with calculated cells per microliter using neg binom and ZI
```{r}
hist(feces_dates2_noNAs$Cells_per_ul)

# try different variables in the ZI formula 
feces_mod3 <- glmmTMB(Cells_per_ul ~ Treatment.x + Wing_mm + fecal_vol + (1|Colony), 
                      data = feces_dates2, 
                      family = nbinom2(),
                      ziformula = ~Treatment.x + Wing_mm + fecal_vol) # all 3 
simulateResiduals(feces_mod3, plot = T) # looks good 

feces_mod3.1 <- glmmTMB(Cells_per_ul ~ Treatment.x + Wing_mm + fecal_vol + (1|Colony), 
                      data = feces_dates2, 
                      family = nbinom2(),
                      ziformula = ~Treatment.x + Wing_mm) # treat and wing 
simulateResiduals(feces_mod3.1, plot = T) # good 

feces_mod3.2 <- glmmTMB(Cells_per_ul ~ Treatment.x + Wing_mm + fecal_vol+ (1|Colony), 
                      data = feces_dates2, 
                      family = nbinom2(),
                      ziformula = ~Treatment.x + fecal_vol) # treat and feces 
simulateResiduals(feces_mod3.2, plot = T) # good

feces_mod3.3 <- glmmTMB(Cells_per_ul ~ Treatment.x + Wing_mm + fecal_vol + (1|Colony), 
                      data = feces_dates2, 
                      family = nbinom2(),
                      ziformula = ~Wing_mm + fecal_vol) # feces and wing 
simulateResiduals(feces_mod3.3, plot = T) # good

AICtab(feces_mod3, feces_mod3.1, feces_mod3.2, feces_mod3.3) 

#              dAIC df
# feces_mod3.3  0.0 9 
# feces_mod3    1.7 10
# feces_mod3.1  6.2 9 
# feces_mod3.2  6.9 9 

Anova(feces_mod3.3)
# Response: Cells_per_ul
#              Chisq Df Pr(>Chisq)  
# Treatment.x 0.0780  1    0.78007  
# Wing_mm     0.0980  1    0.75419  
# fecal_vol   5.5357  1    0.01863 *
```

#### Plot barplot 
```{r}
feces.means <- emmeans(feces_mod3.3, ~ Treatment.x, type = "response")
feces.means.to.plot<-as.data.frame(summary(feces.means))
feces.means.to.plot$upperSE <- feces.means.to.plot$response + feces.means.to.plot$SE
feces.means.to.plot$lowerSE <- feces.means.to.plot$response - feces.means.to.plot$SE
feces.means.to.plot
```

```{r}
feces_barplot<- 
  ggplot(feces.means.to.plot,aes(x=Treatment.x, y=response, ymin=lowerSE, ymax=upperSE)) + 
  theme_classic() +
  geom_bar(stat="identity", aes(fill = Treatment.x), color = "black") +
  scale_fill_manual(values=c("sienna4", "goldenrod1"),
                    labels = c("buckwheat","sunflower")) +
  ylab(expression(paste(italic("Crithidia"), " cells in 1 μL feces"))) +
  xlab(" ") +
  geom_errorbar(width=0.05, position=position_dodge(width = 0.075))+
  theme(text = element_text(size=18),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  scale_x_discrete(labels = c("Buckwheat", "Sunflower")) +
  ylim(0, 2500) + 
  geom_text(x = 1.5, y = 2250, label = expression(paste("Treatment: ", italic("P"), "= 0.78")), size = 5) 

feces_barplot

ggsave(
  "feces_barplot.jpg",
  plot = feces_barplot,
  width = 2000,
  height =1800,
  units = "px",
  dpi = 300
)
```

## Fecal screen presence/absence
### Make presence/absence column

```{r}
# make a new column with 0 or 1 instead of no or yes 
feces_dates2_noNAs$Cells_presence2 <- NA

for(i in 1:length(feces_dates2_noNAs$Cells_presence)){
  if(feces_dates2_noNAs$Cells_presence[i] == "yes") {
    feces_dates2_noNAs$Cells_presence2[i] <- 1
  } else {
    feces_dates2_noNAs$Cells_presence2[i] <- 0
  }
} 
```

### Model presence/absence 

```{r}
feces_PA_mod <- glmer(Cells_presence2 ~ Treatment.x + Wing_mm + fecal_vol + (1|Crithidia_date) + (1|Colony), 
                      family = binomial(link = "logit"), data = feces_dates2_noNAs)

feces_PA_mod2 <- glmer(Cells_presence2 ~ Treatment.x + Wing_mm + fecal_vol + (1|Colony), 
                      family = binomial(link = "logit"), data = feces_dates2_noNAs)

feces_PA_mod3 <- glmer(Cells_presence2 ~ Treatment.x + Wing_mm + fecal_vol + (1|Crithidia_date), 
                      family = binomial(link = "logit"), data = feces_dates2_noNAs)

feces_PA_mod4 <- glm(Cells_presence2 ~ Treatment.x + Wing_mm + fecal_vol, 
                      family = binomial(link = "logit"), data = feces_dates2_noNAs)

AICtab(feces_PA_mod, feces_PA_mod2, feces_PA_mod3, feces_PA_mod4) # 4 best 

simulateResiduals(feces_PA_mod4, plot = T) # looks good 

Anova(feces_PA_mod4)
# Response: Cells_presence2
#                      LR Chisq Df Pr(>Chisq)   
# Treatment              0.2619  1   0.608822   
# Wing_mm                7.1817  1   0.007365 **
# Feces_line_length_mm   6.4816  1   0.010900 * 
```

### Plot presence/absence 

```{r}
feces.PA.means <- emmeans(feces_PA_mod4, ~ Treatment.x, type = "response")
feces.PA.means.to.plot<-as.data.frame(summary(feces.PA.means))
feces.PA.means.to.plot$upperSE <- feces.PA.means.to.plot$prob + feces.PA.means.to.plot$SE
feces.PA.means.to.plot$lowerSE <- feces.PA.means.to.plot$prob - feces.PA.means.to.plot$SE
feces.PA.means.to.plot
```

```{r}
feces_PA_barplot <- 
  ggplot(feces.PA.means.to.plot,aes(x=Treatment.x, y=prob, ymin=lowerSE, ymax=upperSE)) + 
  theme_classic() +
  geom_bar(stat="identity", aes(fill = Treatment.x), color = "black") +
  scale_fill_manual(values=c("sienna4", "goldenrod1"),
                    labels = c("buckwheat","sunflower")) +
  ylab("Probability of cells in 1 μL feces ") +
  xlab(" ") +
  geom_errorbar(width=0.05, position=position_dodge(width = 0.075))+
  theme(text = element_text(size=18),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  scale_x_discrete(labels = c("Buckwheat", "Sunflower")) +
  ylim(0, 1.1) + 
  geom_text(x = 1.5, y = 1, label = expression(paste("Treatment: ", italic("P"), "= 0.61")), size = 5) 

feces_PA_barplot
```

### COMBINE FOUR BARPLOTS FOR FIGURE 2! 

```{r}
fig2 <- ggarrange(feces_barplot, feces_PA_barplot,
                  gut_barplot, gut_PA_barplot,
                  nrow = 2, ncol = 2,
                  labels = c("a", "b", "c", "d"),
                  common.legend = F)

fig2

ggsave(
  "Yost-fig2.jpg",
  plot = fig2,
  width = 3200,
  height =3200,
  units = "px",
  dpi = 300
)
```
